---
title: "Week 5 Practical: Data Wrangling"
subtitle: "Data Frame Ecosystems + Building a Feature Table — PUBH 6854 / PUBH 4201"
date: "Sep 21 & Sep 23, 2026"
output: html_document
---

This one notebook covers both of this week's lecture days: **Part 1** is the
tidyverse side of Day 1's data-frame-ecosystems live demo (the pandas/polars
side lives in `Jupyter/week05_practical.ipynb` — pick whichever language
you're using for Lab 3, you don't need both), and **Part 2** is Day 2's
samples × features × metadata hands-on exercise.

**Data:** `../gene_counts_long.csv`, `../sample_metadata.csv`,
`../gene_metadata.csv` — a small synthetic gene-expression-style dataset
built for this exercise (see `../SOURCE.md`). Deliberately **not** the Lab 3
dataset — different shape, different gene panel, and clean-but-incomplete
rather than messy, since the point here is the pivot/join/readiness
workflow, not regex-style text parsing.

## Setup

If you don't already have `tidyr` (most of you will, from `tidyverse`),
install it once:

```{r install, eval=FALSE}
install.packages("tidyr")
```

```{r setup, message=FALSE}
library(dplyr)
library(tidyr)

counts <- read.csv("../gene_counts_long.csv", stringsAsFactors = FALSE)
samples <- read.csv("../sample_metadata.csv", stringsAsFactors = FALSE)
genes <- read.csv("../gene_metadata.csv", stringsAsFactors = FALSE)

cat(nrow(counts), "count rows,", nrow(samples), "samples,", nrow(genes), "genes\n")
head(counts)
```

## Part 1 — Data frame ecosystems: the same pipeline, three ways

Monday's lecture walked through **filter → select → mutate → group-by/
summarize → join** side by side in tidyverse, pandas, and polars. Below is
the tidyverse half of that, using our actual data instead of the lecture's
toy example. The pandas/polars version of every chunk here has a
line-for-line equivalent in `Jupyter/week05_practical.ipynb` — worth
skimming even if Python isn't your language this semester, just to see how
the same concepts map across ecosystems.

### Filter

Keep only samples in the `treatment` group.

```{r filter}
treatment <- samples %>% filter(treatment_group == "treatment")
cat("tidyverse:", nrow(treatment), "rows\n")
head(treatment, 3)
```

### Select

Keep only the columns we need for a quick site/group summary.

```{r select}
samples %>% select(sample_id, site, treatment_group) %>% head(3)
```

### Mutate

Add a derived column — `log1p` of the raw count, a common transform for
count data before modeling (compresses the right skew from a handful of
very highly expressed genes like `GAPDH`/`ACTB`).

```{r mutate}
counts <- counts %>% mutate(log_count = log1p(raw_count))
head(counts, 3)
```

### Group-by / summarize

Mean raw count per gene, across all samples — a sanity check that
housekeeping genes (`GAPDH`, `ACTB`) really are the highest-expressed, as
`../gene_metadata.csv`'s `gene_class` column claims.

```{r groupby}
summary_tbl <- counts %>%
  group_by(gene_id) %>%
  summarize(mean_count = round(mean(raw_count), 1)) %>%
  arrange(desc(mean_count))
summary_tbl
```

### Join

Attach sample-level metadata (site, treatment group, age) onto the
long-format counts table — this is the join that sets up Part 2's feature
table.

```{r join}
merged <- counts %>% left_join(samples, by = "sample_id")
head(merged, 3)
```

**Checkpoint:** `merged` should have the same number of rows as `counts` (a
left join keeps every count row, even if — hypothetically — a `sample_id`
had no metadata match) and the same number of columns as `counts` +
`samples` minus the shared `sample_id` key. If your row count changed,
that's usually a sign of a many-to-many join somewhere — worth tracking
down rather than ignoring.

## Wide vs. long / tidy data

**Tidy data principle:** one observation per row, one variable per column.
`gene_counts_long.csv` is tidy *for a "count" observation* (each row is one
sample-gene measurement) — but it's the wrong shape for "one row per
sample," which is what most downstream modeling and plotting tools expect.
Getting from one to the other is a **pivot** (long → wide) or its inverse
(wide → long), not a join — no new information is added, only reshaped.

Below: pivot `counts` from long (one row per sample×gene) to wide (one row
per sample, one column per gene) — the samples × features shape Part 2
builds on.

```{r pivot}
wide <- counts %>%
  select(sample_id, gene_id, raw_count) %>%
  pivot_wider(names_from = gene_id, values_from = raw_count)
wide
```

Notice the `NA` cells — those are the ~6% of sample×gene pairs that were
never in `gene_counts_long.csv` to begin with (see `../SOURCE.md`). The
pivot didn't create this missingness, it just made it *visible* as actual
cells instead of absent rows. That's exactly the kind of thing "analytic
data readiness" means catching before you hand a table to a model — see
Part 2.

## Part 2 — Constructing a samples × features × metadata table (Day 2)

**The canonical shape:** one row per sample, one column per feature (here,
genes), plus sample-level metadata columns (age, sex, site,
treatment_group) either alongside the feature columns or joinable by
`sample_id`. This is the shape nearly every downstream analysis —
statistical modeling, ML, visualization — expects.

**Recap of the pieces:**

- `gene_counts_long.csv` — raw pipeline export, long format (what Part 1
  pivoted above)
- `sample_metadata.csv` — one row per sample, clinical/demographic fields
- `gene_metadata.csv` — one row per gene, feature-level annotation

### Your turn: build the full feature table

Using `wide` from above, join in `sample_metadata.csv` so every row has
both the gene columns *and* the sample metadata. Fill in the chunk below.

```{r feature-table}
# TODO: left_join `wide` and `samples` on sample_id
feature_table <- wide %>% left_join(samples, by = "sample_id")
feature_table
```

### Data readiness check

Before this table goes anywhere near a model, run through a short
checklist — this is what "analytic data readiness" means concretely, not
just an abstract idea:

1. **Consistent types** — is every gene column numeric? Is `age` numeric
   (watch for a blank cell silently turning a column into character)?
2. **Documented missingness** — how many missing values, where, and do you
   know *why* (dropped-out gene measurement vs. a metadata field nobody
   filled in are different problems with different fixes)?
3. **No leakage between metadata and features** — `treatment_group` is
   useful to know, but would using it as a *feature* in a model that's
   trying to predict treatment response be circular? (No single right
   answer — the point is to ask.)
4. **Units resolved** — not an issue for raw counts here, but always check
   for a real dataset (see Lab 3's `messy_samples.csv` for what it looks
   like when this *isn't* resolved).

Run the missingness check below, then answer the two questions after it.

```{r missingness}
# TODO: count NAs per column in feature_table
colSums(is.na(feature_table))
```

**Answer these (no need to write code — just think it through, same as
the week's discussion prompt):**

1. Which column(s) have the most missing values, and does that match what
   `../SOURCE.md` told you to expect?
2. `age` is missing for one sample. Before you decide how to handle it
   (drop the row? impute the mean? leave it and let downstream tools
   handle `NA`?) — what would you need to know first, and who would you
   ask? (This is this week's discussion prompt, applied to your own
   table.)

### Optional extension: bring in feature-level metadata

`gene_metadata.csv` has a `gene_class` column (`housekeeping`, `oncogene`,
`cytokine`, ...). Try filtering `feature_table` down to just the cytokine
genes (`IL6`, `TNF`) using `genes` to look up which columns those are — a
small taste of how feature *and* sample metadata both matter when you're
deciding what to include in an analysis.

```{r cytokine-subset}
# TODO (optional): filter genes to gene_class == "cytokine", get the
# gene_id vector, and select just those columns (+ sample_id) from feature_table
cytokine_genes <- genes %>% filter(gene_class == "cytokine") %>% pull(gene_id)
feature_table %>% select(sample_id, all_of(cytokine_genes))
```

## Wrap-up

| | |
|---|---|
| **Due Wed, Sep 30, 11:59pm** | Lab 3: Parsing Messy Health or Genomic Data |
| **Keep for later** | This "samples × features × metadata" shape comes back directly in Week 8-9 (Lab 4: Database-Driven Feature Table) — same framing, different (relational-database) source data |
| **Closes Module 1** | Environments (Wk1-2) → notebooks (Wk3) → text parsing (Wk4) → structured tables (Wk5) — Module 2 (Wk6+) shifts to software design |
