---
title: "Week 4 Practical: Text Processing"
subtitle: "Regex + AI-Assisted Extraction — PUBH 6854 / PUBH 4201"
date: "Wednesday, Sep 16, 2026"
output: html_document
---

Monday's lecture gave AI-assisted extraction a 5-minute preview on purpose —
this is where it gets real time. Today: regex hands-on, then a proper deep
dive into AI-assisted extraction and how to troubleshoot with AI, then a
head-to-head comparison using the same data.

**Data:** `../inclass_headers.fasta` — 9 messy FASTA headers, each with
sample ID, organism, and gene fields, formatted inconsistently on purpose:
pipe-delimited, colon-delimited, semicolon-delimited, and space-separated
variants; several headers missing a length field entirely (one explicitly
marked `len:NA`); and one hyphenated gene symbol (`HLA-DRB1`). This is a
different, smaller set than Lab 3's dataset — today's answers won't just
transfer over.

## Setup

Minimal setup on purpose — if you don't already have `stringr` (e.g. from
Week 2's `renv` work, or as part of `tidyverse`), install it once:

```{r install, eval=FALSE}
install.packages("stringr")
```

```{r setup, message=FALSE}
library(stringr)

headers <- readLines("../inclass_headers.fasta")
headers <- headers[str_starts(headers, ">")]
headers
```

## Part 1 — Regex extraction (20 min)

Working in pairs, write regex patterns to extract three fields from every
header:

1. **Sample ID** (e.g. `seq101`, `Seq102`, `seq-103` — note the inconsistent
   casing and punctuation)
2. **Organism** (mostly `Homo_sapiens` / `Homo sapiens` / `Hsapiens` —
   another inconsistency to handle, or explicitly decide to ignore)
3. **Gene** (`KRAS`, `PTEN`, `TP53`, and one `HLA-DRB1` — the hyphen from
   Monday's live-code lesson shows up for real here)

The `gene` pattern below is filled in already — it's Monday's live-code
pattern, extended to handle both `gene=` and `gene:`. You'll need to write
your own patterns for `sample_id` and `organism`.

**Test against all nine headers, not just the first one or two** — several
are deliberately built to break a pattern that only handles the "clean"
cases.

```{r part1-patterns}
sample_id_pattern <- "TODO"                 # fill this in
organism_pattern  <- "TODO"                 # fill this in
gene_pattern      <- "(?i)gene[:=]([\\w-]+)" # Monday's pattern, extended for : and gene: (case-insensitive)

extracted <- data.frame(
  header    = headers,
  sample_id = str_extract(headers, sample_id_pattern),
  organism  = str_extract(headers, organism_pattern),
  gene      = str_match(headers, gene_pattern)[, 2],
  stringsAsFactors = FALSE
)
extracted
```

**Checkpoint:** by the end of Part 1, you should have three working
patterns and a quick sanity check of each against the full table above —
not full production code, just confidence that they hold up across all
nine rows. No `NA`s in the `gene` column is a good sign; `NA`s in
`sample_id` or `organism` mean your pattern doesn't handle every format
yet.

## Part 2 — Deep dive: AI-assisted extraction & troubleshooting with AI (25 min)

This section is discussion-driven, not code-driven — your instructor will
walk through this live. A short summary to have in front of you:

### Writing a real extraction prompt

A vague prompt ("extract the fields from this") gets a vague, inconsistent
answer. A good extraction prompt is specific about:

- **What fields you want, named explicitly** ("sample ID, organism, and
  gene symbol")
- **The exact input**, pasted in full — not summarized or retyped
- **The output format you want** (a table, a list, CSV) so you can
  actually compare it against your regex output
- **What to do with a missing field** (leave blank? mark `NA`? — decide
  this yourself, don't let the AI tool decide silently)

### The chat-vs-agentic distinction

Some AI tools just **explain** — they describe what a header contains and
you still do the extraction. Others can **act** — actually run code and
produce the extraction for you. Neither is strictly better, but you need
to know which one you're using: an explaining tool is a starting point
for your own regex or script; an acting tool is producing output you need
to verify, not just accept.

### The verification habit

This is the part that matters most, because **AI extraction fails
differently than regex does.** A broken regex either throws an error or
visibly extracts the wrong thing (like the `HLA` truncation from Monday's
demo — wrong, but noticeably short). AI extraction can fail with a
completely confident, well-formatted, wrong answer — a hallucinated field
that was never in the source text, or a "cleaned up" value that no longer
matches what was actually written. There is no error message for this.

**The habit:** for every field an AI tool extracts, spot-check it against
the actual source text before you trust it.

### Try it yourself

Paste your extraction prompt and the AI's response here for one header, as
a quick note to yourself — not graded, just so you have a record to refer
back to in Part 3.

> **Prompt:**
>
> **AI's answer:**

## Part 3 — Hands-on: same task, via AI (15 min)

Using the prompting approach from Part 2, run all nine headers through an
AI tool and extract the same three fields (sample ID, organism, gene).
Fill in what it gives you below, then compare against your Part 1 regex
output.

```{r part3-ai-results}
ai_results <- data.frame(
  header    = headers,
  sample_id = NA_character_,
  organism  = NA_character_,
  gene      = NA_character_,
  stringsAsFactors = FALSE
)
# Fill in by hand from your AI tool's output, e.g.:
# ai_results$gene <- c("KRAS", "PTEN", "TP53", "KRAS", "PTEN", "TP53", "KRAS", "PTEN", "HLA-DRB1")
ai_results
```

```{r part3-compare}
comparison <- data.frame(
  header     = headers,
  regex_gene = extracted$gene,
  ai_gene    = ai_results$gene,
  agree      = extracted$gene == ai_results$gene,
  stringsAsFactors = FALSE
)
comparison
```

For each of the 27 field values (9 headers × 3 fields):

- Do they agree?
- If not, which one is actually right? (Go back to the source text — this
  is the verification habit from Part 2, applied for real.)
- Is there a pattern to where they disagree? (Hint: check the `HLA-DRB1`
  row and the rows missing a length field particularly closely.)

## Discussion (5 min)

When would you reach for regex vs. AI on a new extraction task you haven't
seen before? There's no single right answer — the point is to leave with a
real basis for that decision, not just "AI is easier" or "regex is more
reliable" as a slogan.

## Wrap-up

| | |
|---|---|
| **Due tonight, 11:59pm** | Lab 2: Analysis Notebook |
| **Next week** | Lab 3: Parsing Messy Health or Genomic Data (assigned Week 5, due Sep 30) — builds directly on today, with a new dataset |
| **Keep for later** | The prompting checklist and troubleshoot-with-AI steps above apply well beyond text extraction — worth remembering for Week 7 (AI-assisted debugging) and Week 8 (AI-assisted SQL) |
