---
title: "Week 3 Practical -- Starter Notebook (R Markdown)"
output: html_document
---

This notebook has **one bug**. It was "tested" by running chunks out of
order in the RStudio console while writing it, so it happened to work in
that session -- but the order on the page doesn't match the order the
chunks were actually run in.

**Your task:**

1. Click **Knit** (or run `rmarkdown::render("starter_notebook.Rmd")`)
   and watch it fail.
2. Diagnose the error with an AI assistant (paste the exact error text).
3. Fix it, then **Knit** again until it completes clean, from a fresh R
   session.
4. Add one new Markdown section and one new R chunk of your own below the
   last chunk.
5. Knit one more time to confirm your addition doesn't break anything.
6. The rendered `starter_notebook.html` this produces *is* your export --
   nothing extra to run for that step.

## Summarize by site

```{r summarize-bug}
patients %>%
  group_by(site) %>%
  summarize(mean_age = mean(age))
```

## Setup

(this belongs above the chunk that uses it -- that's the bug)

```{r setup}
library(dplyr)

patients <- data.frame(
  patient_id = c("P001", "P002", "P003", "P004", "P005"),
  age = c(54, 61, 47, 72, 39),
  site = c("DC", "MD", "DC", "VA", "MD")
)
```

## Count by site

```{r count-by-site}
patients %>%
  group_by(site) %>%
  summarize(n = n())
```
